---
title: "Project 1"
author: "Nawa Raj"
date: "June 13, 2016"
output: html_document
---
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```

##Project Introduction
##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#Data

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

#Goal

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

#Getting and loading data

```{r}
rm(list = ls())
if (!file.exists("pml-training.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")
}
submit <- read.csv("pml-testing.csv", sep = ",", na.strings = c("", "NA"))
data <- read.csv("pml-training.csv", sep = ",", na.strings = c("", "NA"))

```

#cleanup the data

Here, I remove columns full of NAs and remove features that are not in the submit set. The features containing NAs are the variance, mean and stddev within each window for each feature. Since the submit dataset has no time-dependence, these values are useless and can be disregarded. I also remove the first 7 features since they are related to the time-series or are not numeric.


```{r}
# Remove columns full of NAs.
features <- names(submit[,colSums(is.na(submit)) == 0])[8:59]
# Only use features used in submit cases.
data <- data[,c(features,"classe")]
submit <- submit[,c(features,"problem_id")]
```


#Bootstrap

Next, I withhold 25% of the dataset for testing after the final model is constructed

```{r}
set.seed(916)
inTrain = createDataPartition(data$classe, p = 0.75, list = F)
training = data[inTrain,]
testing = data[-inTrain,]

```

#Feature Selection

Some features may be highly correlated. The PCA method mixes the final features into components that are difficult to interpret; instead, I drop features with high correlation (>90%).

```{r}
outcome = which(names(training) == "classe")
highCorrCols = findCorrelation(abs(cor(training[,-outcome])),0.90)
highCorrFeatures = names(training)[highCorrCols]
training = training[,-highCorrCols]
outcome = which(names(training) == "classe")
```
 
The features with high correlation are accel_belt_z, roll_belt, accel_belt_y, accel_belt_x, gyros_arm_y, gyros_forearm_z, and gyros_dumbbell_x.  

#Feature Importance

The random forest method reduces overfitting and is good for nonlinear features. First, to see if the data is nonlinear, I use the random forest to discover the most important features. The feature plot for the 4 most important features is shown.
 
```{r}
fsRF = randomForest(training[,-outcome], training[,outcome], importance = T)
rfImp = data.frame(fsRF$importance)
impFeatures = order(-rfImp$MeanDecreaseGini)
inImp = createDataPartition(data$classe, p = 0.05, list = F)
featurePlot(training[inImp,impFeatures[1:4]],training$classe[inImp], plot = "pairs")

 
```
THe most important features are :
**pitch_belt**
**yaw_belt**
**total_accel_belt**
**gyros_belt_x**


##Training

Use random forest ans k nearest neighbours algorithms.
```{r}
ctrlKNN = trainControl(method = "adaptive_cv")
modelKNN = train(classe ~ ., training, method = "knn", trControl = ctrlKNN)
ctrlRF = trainControl(method = "oob")
modelRF = train(classe ~ ., training, method = "rf", ntree = 200, trControl = ctrlRF)
resultsKNN = data.frame(modelKNN$results)
resultsRF = data.frame(modelRF$results)

```

#Testing out of sample error
The random forest will give a larger accuracy compared to k-nearest neighbors. Here, I give the confusion matrix between the KNN and RF models to see how much they agree on the test set, then I compare each model using the test set outcomes.


```{r}
fitKNN = predict(modelKNN, testing)
fitRF = predict(modelRF, testing)
```
#KNN Vs RF

I have copied and paste the output Required output only.



Confusion Matrix and Statistics

#           Reference
# Prediction    A    B    C    D    E
#          A 1360    8   12   16    7
#          B   37  834   27   24   25
#          C   14   31  780   22    9
#          D    9    3   56  721   14
##          E   14   33   22   36  790

## Overall Statistics
                                       
##                Accuracy : 0.915         
##                  95% CI : (0.906, 0.922)
##     No Information Rate : 0.292         
##     P-Value [Acc > NIR] : < 2e-16       
                                       
##                   Kappa : 0.892         
##  Mcnemar's Test P-Value : 3.1e-11       

## Statistics by Class:

##                      Class: A Class: B Class: C Class: D Class: E
## Sensitivity             0.948    0.917    0.870    0.880    0.935
## Specificity             0.988    0.972    0.981    0.980    0.974
## Pos Pred Value          0.969    0.881    0.911    0.898    0.883
## Neg Pred Value          0.979    0.981    0.971    0.976    0.986
## Prevalence              0.292    0.185    0.183    0.167    0.172
## Detection Rate          0.277    0.170    0.159    0.147    0.161
## Detection Prevalence    0.286    0.193    0.175    0.164    0.183
## Balanced Accuracy       0.968    0.945    0.925    0.930    0.955

#KNN VS TEST

## Confusion Matrix and Statistics

##           Reference
## Prediction    A    B    C    D    E
##          A 1354   44   13    9   14
##          B    6  835   32    3   33
##          C   12   25  779   59   22
##          D   16   24   20  723   36
##          E    7   21   11   10  796

## Overall Statistics
                                        
##                Accuracy : 0.915         
##                  95% CI : (0.907, 0.923)
##     No Information Rate : 0.284         
##     P-Value [Acc > NIR] : < 2e-16       
##                                         
##                   Kappa : 0.892         
##  Mcnemar's Test P-Value : 3.91e-15      
## 
## Statistics by Class:
## 
##                      Class: A Class: B Class: C Class: D Class: E
## Sensitivity             0.971    0.880    0.911    0.899    0.883
## Specificity             0.977    0.981    0.971    0.977    0.988
## Pos Pred Value          0.944    0.919    0.868    0.883    0.942
## Neg Pred Value          0.988    0.971    0.981    0.980    0.974
## Prevalence              0.284    0.194    0.174    0.164    0.184
## Detection Rate          0.276    0.170    0.159    0.147    0.162
## Detection Prevalence    0.292    0.185    0.183    0.167    0.172
## Balanced Accuracy       0.974    0.931    0.941    0.938    0.936
#RFvs. test set

## Confusion Matrix and Statistics
## 
##           Reference
## Prediction    A    B    C    D    E
##          A 1395    8    0    0    0
##          B    0  940    7    0    0
##          C    0    1  845   10    0
##          D    0    0    3  794    6
##          E    0    0    0    0  895
## 
## Overall Statistics
##                                        
##                Accuracy : 0.993        
##                  95% CI : (0.99, 0.995)
##     No Information Rate : 0.284        
##     P-Value [Acc > NIR] : <2e-16       
##                                        
##                   Kappa : 0.991        
##  Mcnemar's Test P-Value : NA           
## 
## Statistics by Class:
## 
##                      Class: A Class: B Class: C Class: D Class: E
## Sensitivity             1.000    0.991    0.988    0.988    0.993
## Specificity             0.998    0.998    0.997    0.998    1.000
## Pos Pred Value          0.994    0.993    0.987    0.989    1.000
## Neg Pred Value          1.000    0.998    0.998    0.998    0.999
## Prevalence              0.284    0.194    0.174    0.164    0.184
## Detection Rate          0.284    0.192    0.172    0.162    0.183
## Detection Prevalence    0.286    0.193    0.175    0.164    0.183
## Balanced Accuracy       0.999    0.994    0.993    0.993    0.997

#Conclucion

The random forest fit is clearly more accurate than the k-nearest neighbors method with 99% accuracy.


